---
title: "PA1_template.Rmd"
author: "Joe"
date: "Sunday, October 18, 2015"
output: html_document
---

```{r readdata}
## Read the data
actData <- read.csv("activity.csv")
```

***

## Steps taken per day

```{r stepsperday}
## Sum of steps per day
sumSteps <- aggregate(actData$steps, list(actData$date), sum)
## Create histogram
hist(sumSteps$x, col = "pink", main = "Histogram of total steps taken per day", xlab = "Total steps taken per day")
## Calculate mean and median
meanSteps <- as.integer(mean(sumSteps$x, na.rm=TRUE))
medianSteps <- median(sumSteps$x, na.rm=TRUE)
```

The mean number of steps per day is **`r meanSteps`**.  
The median number of steps per day is **`r medianSteps`**.

***

## Average daily activity pattern

```{r dailyactivity}
## Average per interval
meanInterval <- aggregate(actData$steps, list(actData$interval), mean, na.rm=TRUE)
## Create time series chart
plot(meanInterval$Group.1, meanInterval$x, xlab = "Interval",
     ylab = "Average steps", main = "Average daily activity pattern", type="n")
lines(meanInterval$Group.1, meanInterval$x)
## Calculate maximum
maxInterval <- meanInterval$Group.1[order(meanInterval$x, decreasing=T)[1]]
```

The 5-minute interval with the largest number of average steps per day is
interval **`r maxInterval`**.  

***

## Imputing missing values

```{r countmissing}
## Count missing values
countNA <- sum(is.na(actData$steps))
```

There are **`r countNA`** total missing values in the dataset.  
To impute values, we'll simply replace these NAs with the mean value for their interval.

```{r impute}
## Merge original data with interval means
actData2 <- merge(actData, meanInterval, by.x = "interval", by.y = "Group.1")
## Use interval means as replacements for NAs
actData2$steps[is.na(actData2$steps)] <- actData2$x[is.na(actData2$steps)]
## Sum of steps per day
sumSteps2 <- aggregate(actData2$steps, list(actData2$date), sum)
## Create histogram
hist(sumSteps2$x, col = "green", main = "Histogram of total steps taken per day", xlab = "Total steps taken per day")
## Calculate mean and median
meanSteps2 <- as.integer(mean(sumSteps2$x, na.rm=TRUE))
medianSteps2 <- as.integer(median(sumSteps2$x, na.rm=TRUE))
```

The mean number of steps per day is now **`r meanSteps2`** *(previous value `r meanSteps`)*.  
The median number of steps per day is now **`r medianSteps2`** *(previous value `r medianSteps`)*.  
There is very little difference in mean and median; however, unsurprisingly, the total steps have increased (see y axes of histograms).

***

## Weekdays vs weekends

```{r weekday}
## Establish which days are weekdays and weekends
actData2$date <- as.Date(actData2$date)
actData2$weekday <- weekdays(actData2$date)
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
actData2$wDay <- factor((actData2$weekday %in% weekdays1), levels=c(FALSE, TRUE),
                        labels=c('weekend', 'weekday'))
## Split data files
actWD <- subset(actData2, wDay == "weekday")
meanIntervalWD <- aggregate(actWD$steps, list(actWD$interval), mean, na.rm=TRUE)
actWE <- subset(actData2, wDay == "weekend")
meanIntervalWE <- aggregate(actWE$steps, list(actWE$interval), mean, na.rm=TRUE)
## Create panel plot
par(mfrow=c(2,1))
plot(meanIntervalWD$Group.1, meanIntervalWD$x, xlab = "Interval",
     ylab = "Average steps", main = "Average weekday activity pattern", type="n")
lines(meanIntervalWD$Group.1, meanIntervalWD$x)
plot(meanIntervalWE$Group.1, meanIntervalWE$x, xlab = "Interval",
     ylab = "Average steps", main = "Average weekend activity pattern", type="n")
lines(meanIntervalWE$Group.1, meanIntervalWE$x)
```
